---
title: "introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(gimmefMRI)
```

    # Example Model
install the ""gimmefMRI"" package in R using the following two commands:"
install.packages('devtools')
devtools::install_github('jbartolotti/gimmefMRI')

To run, load the gimmefMRI library, and run the program using the following two commands:
library('gimmefMRI')
gimmefMRI()
You will be prompted to select a file. Select a configuration excel file, and the program will run.

For an example with sample data, run 
gimmefMRI('example')

    # Configuration
All data to be analyzed is located in the TIMECOURSES sheet. This contains a single column for each ROI or other predictor (e.g., task) of interest. Models may use all or a subset of these predictors. Additional columns specify Subject, Subgroup, Run, Condition, Slice Number, and Time. The Censor column can be used to exclude single rows from the model (1 = exclude). 

The CONTROL sheet specifies where data is stored and results should be saved, and which parts of the analysis to run.

The MODELS sheet contains any number of columns, each one specifying a single GIMME model to run.

The LISTS sheet contains lists of nodes or subjects to include in a GIMME model, and is referred to within the MODELS configuration.

The ABBREVIATIONS sheet provides a mapping between long names and shortnames for network nodes. Longnames refer to column names in TIMECOURSES that are specified in the LISTS sheet. Shortnames are used in figures.

    ## Runtime options
gimmefMRI('load', 
  run = c(generate_models = TRUE, run_models = TRUE, generate_figures = TRUE, run_figures = TRUE),
  models = c('model1','model2')
  )

Use the `run` option to specify steps to run or skip. Overrides values in the CONTROL sheet.
Use the `models` option to specify models to run for each step. Default: run all models listed in the MODELS sheet.

